Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-01-29 doc/invt.tex
add asymptotic error discussion and errcomp.gp

From Eq. \eqref{eq:alpha_opt},
it is reasonable to conjecture that
a near-optimal $\alpha(t)$ for
a given updating matrix $\mathbf w$ would be
%
\begin{equation}
\alpha(\tau) = \frac{1}{\lambda \, (\tau + t_0) },
\label{eq:alpha_trial}
\end{equation}
%
for some value of $\lambda$.
%
Below we shall find the optimal value
in some special cases.
%
Using Eq. \eqref{eq:alpha_trial}
in Eq. \eqref{eq:error_asym1} yields
%
\begin{align}
\Err_A
&=
\frac{    1    }
     { t + t_0 }
\sum_{k = 2}^n
  \frac{ \Gamma_k \, \nu_k^2 }
       {    2 \, \nu_k - 1   }
\left[
  1 - \left(
        1 + \frac{ t }{ t_0 }
      \right)^{1 - 2 \, \nu_k}
\right],
\label{eq:error_asym_invt}
\end{align}
%
where $\nu_k \equiv \lambda_k / \lambda$.
%
At long times, we get
$$
\begin{aligned}
  \Err_A
  \stackrel{         t \to \infty         }
           { =\joinrel=\joinrel=\joinrel= }
  \sum_{k = 2}^n
  \begin{dcases}
    \frac{    1    }
         { t + t_0 }
    \frac{ \Gamma_k \, \nu_k^2 }
         {   2 \, \nu_k - 1    }
    &
    \mathrm{if \;} 2 \, \nu_k > 1,
    \\%[1em]
    %
    %
    \frac{    \Gamma_k    }
         { 4 \, (t + t_0) }
    \ln \frac{ t + t_0 }
             {   t_0  }
    &
    \mathrm{if \;} 2 \, \nu_k = 1,
    \\
    %
    %
    \frac{  t_0^{ 2 \, \nu_k  - 1}  }
         { (t + t_0)^{ 2 \, \nu_k } }
    \frac{ \Gamma_k \, \nu_k^2 }
         {   1 - 2 \, \nu_k    }
    &
    \mathrm{if \;} 2 \, \nu_k < 1.
  \end{dcases}
\end{aligned}
$$
%
The last two cases, $\lambda \ge 2\lambda_k$,
are asymptotically suboptimal
as they decay slower than $1/t$.
%
Thus, the optimal $\lambda$
satisfies $\lambda < 2 \min \lambda_k$,
and in this case,
we can seek that
the optimal value from
the smallest real root of
%
\begin{equation}
\sum_{k = 2}^n
\frac{ \lambda_k - \lambda }
{ \left(2 - \lambda/ \lambda_k \right)^2 }
= 0.
\label{eq:optimal_lambda_approx}
\end{equation}

@@

In practice, however, we may wish to optimize
relative short simulations that have not entered
the asymptotic regime.
%
Then the residual error, $\Err_R$, is not negligible.
%
From Eq. \eqref{eq:error_res}, we get
%
\begin{equation}
\Err_R
=
\sum_{k = 2}^n
  \frac{ \Gamma_k \, \nu_k }
       {        2 \, t_0   }
  \left(
      \frac{   t_0   }
           { t + t_0 }
   \right)^{ 2 \, \nu_k }
\label{eq:error_res_invt}
\end{equation}
%
where we have assumed that at $t = 0$
the system has achieved an equilibrium
at a constant $\alpha_0$,
such that the values of
$\left\langle y_k^2(0) \right\rangle$
can be computed from Eq. \eqref{eq:y2_eql}.
%
The total error is given by the sum of
Eqs. \eqref{eq:error_res_invt} and \eqref{eq:error_asym_invt}.
%
\begin{align}
\Err
&=
\Err_R + \Err_A
\notag
\\
&=
\sum_{ k = 2 }^n
  \Err^f_k \,
  \left\{
    1
    +
    \frac{ 1 - \left[ t_0 / (t + t_0) \right]^{2 \, \nu_k - 1} }
         {                 2 \, \nu_k - 1                      }
%    \frac{       1        }
%         { 2 \, \nu_k - 1 }
%    \left[
%      1 -
%      \left(
%        \tfrac{   t_0   }
%              { t + t_0 }
%      \right)^{2 \, \nu_k - 1}
%    \right]
  \right\}
,
\label{eq:error_invt}
\end{align}
%
where
$$
\Err^f_k
\equiv
\left\langle
  y_k^2
\right\rangle_{ \alpha(t) }
=
\frac{  \Gamma_k \, \nu_k   }
     {    2 \, (t + t_0)    }
,
$$
is the equilibrium error of mode $k$
under the constant terminal value
of $\alpha(t)$.

@@

We give two remarks on Eq. \eqref{eq:error_invt}.
%
First, the final error is always greater than
the saturated equilibrium value at the terminal $\alpha(t)$:
$$
E > E_\mathrm{sat} \equiv \sum_{k = 2}^n E^f_k,
$$
because the second term in the braces
is nonnegative no matter the sign of $2 \, \nu_k - 1$.
%
Second, in the optimal case,
$\nu_k = 1$,
and the final error is twice $E_\mathrm{sat}$
at long times.

@@

The Gaussian kernel is commonly
used in metadynamics.
%
Below we show that this
leads to a stable updating scheme,
although it can be less effective
in reducing local noises.

We start by defining
a continuous limit
for $b = n - 1 \gg 1$:
$x = l \, \Delta x$,
$\Delta x = \pi/n$,
and
$\mu_l = \mu(x) \, \Delta x$.
%
Then
Eq. \eqref{eq:wband_eigenvalue}
can be approximated by an integral:
%
\begin{equation}
\lambda_k
=
2 \int_0^\pi
  \mu(x) \, \cos \bigl( (k-1) \, x \bigr) \, dx,
\label{eq:lambda_int}
\end{equation}
%
with the normalization
%
\begin{equation}
1 = 2 \int_0^\pi \mu(x) \, dx.
\label{eq:mx_normalization}
\end{equation}

For the Gaussian wave,
$\mu(x) \propto \exp(-\frac{1}{2} x^2/\sigma^2)$.
%
If $\Delta x \ll \sigma \ll 1$,
we can extended
the upper limit of the integrals
in Eqs. \eqref{eq:lambda_int}
and \eqref{eq:mx_normalization}
to infinity, and
%
\begin{equation}
\mu(x)
=
\frac{            1            }
     { \sqrt{ 2 \pi \sigma^2 } }
%
\exp\left(
      -\frac{       x^2     }
            { 2 \, \sigma^2 }
    \right),
\end{equation}
%
%
The eigenvalues are given by
%
\begin{align}
\lambda_k
&=
\exp\left[
      -\frac{ (k - 1)^2 \, \sigma^2 }
            {           2           }
    \right]
\notag
\\
&=
\exp\left[
      -\frac{ \pi^2 }{ 2 }
      \left(
        \frac{ k - 1 }
             {   n   }
      \right)^2
      \left(
        \frac{  \sigma }
             { \Delta x }
      \right)^2
    \right]
.
\label{eq:lambda_Gaussian}
\end{align}
%
The eigenvalues are all positive,
suggesting stability of the updating scheme.

>>>>

For an arbitrary updating matrix $\mathbf w$,
optimizing Eq. \eqref{eq:error_asym1}
can lead to an algebraically complex expression for $\alpha(t)$.
%
In practice, it is more useful to find a simple expression for $\alpha(t)$
that is near optimal.
%
From Eq. \eqref{eq:alpha_opt},
a reasonable candidate would be
%
\begin{equation}
\alpha(\tau) = \frac{1}{\lambda \, (\tau + t_0) },
\label{eq:alpha_trial}
\end{equation}
%
for some value of $\lambda$.
%
Below we shall assume this functional form
and find the optimal value
in some cases.
%
We shall show that the optimal $\lambda$
generally depends on not only the spectrum
of eigenvalues and the sampling method,
but also other parameters,
such as the simulation length, $t$.

Using Eq. \eqref{eq:alpha_trial}
in Eq. \eqref{eq:error_asym1} yields
%
\begin{align}
\Err_A
&=
\frac{    1    }
     { t + t_0 }
\sum_{k = 2}^n
  \frac{ \Gamma_k \, \nu_k^2 }
       {    2 \, \nu_k - 1   }
\left[
  1 - \left(
        1 + \frac{ t }{ t_0 }
      \right)^{1 - 2 \, \nu_k}
\right],
\label{eq:error_asym_invt}
\end{align}
%
where $\nu_k \equiv \lambda_k / \lambda$.
%
At long times, we get
$$
\begin{aligned}
  \Err_A
  \stackrel{         t \to \infty         }
           { =\joinrel=\joinrel=\joinrel= }
  \sum_{k = 2}^n
  \begin{dcases}
    \frac{    1    }
         { t + t_0 }
    \frac{ \Gamma_k \, \nu_k^2 }
         {   2 \, \nu_k - 1    }
    &
    \mathrm{if \;} 2 \, \nu_k > 1,
    \\%[1em]
    %
    %
    \frac{    \Gamma_k    }
         { 4 \, (t + t_0) }
    \ln \frac{ t + t_0 }
             {   t_0  }
    &
    \mathrm{if \;} 2 \, \nu_k = 1,
    \\
    %
    %
    \frac{  t_0^{ 2 \, \nu_k  - 1}  }
         { (t + t_0)^{ 2 \, \nu_k } }
    \frac{ \Gamma_k \, \nu_k^2 }
         {   1 - 2 \, \nu_k    }
    &
    \mathrm{if \;} 2 \, \nu_k < 1.
  \end{dcases}
\end{aligned}
$$
%
The last two cases, $\lambda \ge 2\lambda_k$,
are asymptotically suboptimal
as they decay slower than $1/t$.
%
Thus, the optimal $\lambda$
satisfies $\lambda < 2 \min \lambda_k$,
and in this case,
we can seek that
the optimal value from
the smallest real root of
%
\begin{equation}
\sum_{k = 2}^n
\frac{ \lambda_k - \lambda }
{ \left(2 - \lambda/ \lambda_k \right)^2 }
= 0.
\label{eq:optimal_lambda_approx}
\end{equation}

@@

In practice, however,
real simulations
may be relative short
and have not entered
the asymptotic regime.
%
Then, the residual error, $\Err_R$, is not negligible.
%
From Eq. \eqref{eq:error_res}, we get
%
\begin{equation}
\Err_R
=
\sum_{k = 2}^n
  \frac{ \Gamma_k \, \nu_k }
       {        2 \, t_0   }
  \left(
      \frac{   t_0   }
           { t + t_0 }
   \right)^{ 2 \, \nu_k },
\label{eq:error_res_invt}
\end{equation}
%
assuming that at $t = 0$
the system has achieved an equilibrium
at a constant $\alpha_0$,
such that the values of
$\left\langle y_k^2(0) \right\rangle$
can be computed from Eq. \eqref{eq:y2_eql}.
%
The total error is given by the sum of
Eqs. \eqref{eq:error_res_invt} and \eqref{eq:error_asym_invt}.
%
\begin{align}
\Err
&=
\Err_R + \Err_A
\notag
\\
&=
\sum_{ k = 2 }^n
  \Err^f_k \,
  \left\{
    1
    +
    \frac{ 1 - \left[ t_0 / (t + t_0) \right]^{2 \, \nu_k - 1} }
         {                 2 \, \nu_k - 1                      }
%    \frac{       1        }
%         { 2 \, \nu_k - 1 }
%    \left[
%      1 -
%      \left(
%        \tfrac{   t_0   }
%              { t + t_0 }
%      \right)^{2 \, \nu_k - 1}
%    \right]
  \right\}
,
\label{eq:error_invt}
\end{align}
%
where
$$
\Err^f_k
\equiv
\left\langle
  y_k^2
\right\rangle_{ \alpha(t) }
=
\frac{  \Gamma_k \, \nu_k   }
     {    2 \, (t + t_0)    }
,
$$
is the equilibrium error of mode $k$
under the constant terminal value
of $\alpha(t) = 1/[\lambda (t + t_0)]$.

@@

We give several remarks on Eq. \eqref{eq:error_invt}.
%
First, the final error is always greater than
the saturated equilibrium value at the terminal $\alpha(t)$:
$$
E > E_\mathrm{sat} \equiv \sum_{k = 2}^n E^f_k,
$$
because the second term in the braces
is nonnegative no matter the sign of $2 \, \nu_k - 1$
(in fact, it is a decreasing function
that vanishes only at $\nu_k \to +\infty$).

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/errcomp.pdf}
  }
  \caption{
    \label{fig:err_component}
    Component of the error, $G(\nu_k, r)$,
    versus $\nu_k = \lambda_k / \lambda$,
    where $r = t_0 / (t + t_0)$.
    %
  }
\end{center}
\end{figure}

Second, we shall show that the optimal $\lambda$
depends on the simulation length,
and it tends to decrease as simulation lengthens.
%
Let us first rewrite Eq. \eqref{eq:error_invt} as
%
$$
\Err
=
\sum_{ k = 2 }^n
  \frac{    \Gamma_k    }
       { 2 \, (t + t_0) }
  G\left( \nu_k, \frac{ t_0 } { t + t_0} \right),
$$
%
where the $k$th component is proportional to
%
$$
G\left( \nu_k, r \right)
=
\nu_k
\,
\left(
  1
  +
  \frac{ 1 - r^{2 \, \nu_k - 1} }
       {   2 \, \nu_k - 1       }
\right).
$$
%
As shown in Fig. \ref{fig:err_component},
the function $G(\nu_k, r)$ depends nontrivial on $\nu_k$.
%
For a fixed and sufficiently small $r = t_0 / (t + t_0)$,
$G(\nu_k, r)$ has two local minima along $\nu_k$,
one at $\nu_k = 0$, and the other around $\nu_k = 1$,
with the former being deeper albeit narrower.
%
This shows that for a relatively short simulation,
hence relative large value of $r$,
the minimal error may be achieved at $\nu_k \approx 0$,
corresponding to a large value of $\lambda$
($\lambda \gg \lambda_k$).
%
However, this minimum is unstable as simulation lengthens,
for $G(\nu_k, r)$ increases rapidly
with decreasing $r$ under a fixed $\nu_k$.
%
In other words,
asymptotically, the minimal error is always achieved at
$\nu_k = 1$ or $\lambda = \lambda_k$.
%
In this case,
$G(\nu_k, r) \to G(1, 0) = 2$
and the total error is roughly twice $E_\mathrm{sat}$.

@@

The Gaussian updating kernel is commonly
adopted in metadynamics.
%
Below we show that the resulting updating scheme
is stable, albeit less effective
in reducing short-wavelength noises.

We start by defining
a continuous limit
for $b = n - 1 \gg 1$:
$x = l \, \Delta x$,
and
$\mu_l = \mu(x) \, \Delta x$,
where
$\Delta x = \pi/n$
is the bin size.
%
Then
Eq. \eqref{eq:wband_eigenvalue}
can be approximated by an integral:
%
\begin{equation}
\lambda_k
=
2 \int_0^\pi
  \mu(x) \, \cos \bigl( (k-1) \, x \bigr) \, dx,
\label{eq:lambda_int}
\end{equation}
%
with the normalization
%
\begin{equation}
1 = 2 \int_0^\pi \mu(x) \, dx.
\label{eq:mx_normalization}
\end{equation}

Particularly,
for the Gaussian updating kernel,
if $\Delta x \ll \sigma \ll 1$,
we can extended
the upper limit of the integrals
in Eqs. \eqref{eq:lambda_int}
and \eqref{eq:mx_normalization}
to infinity, and
%
\begin{equation}
\mu(x)
=
\frac{            1            }
     { \sqrt{ 2 \pi \sigma^2 } }
%
\exp\left(
      -\frac{       x^2     }
            { 2 \, \sigma^2 }
    \right),
\end{equation}
%
%
The eigenvalues are given by
%
\begin{align}
\lambda_k
&=
\exp\left[
      -\frac{ (k - 1)^2 \, \sigma^2 }
            {           2           }
    \right]
\notag
\\
&=
\exp\left[
      -\frac{ \pi^2 }{ 2 }
      \left(
        \frac{ k - 1 }
             {   n   }
      \right)^2
      \left(
        \frac{  \sigma }
             { \Delta x }
      \right)^2
    \right]
.
\label{eq:lambda_Gaussian}
\end{align}
%
The eigenvalues are all positive,
suggesting stability of the updating scheme.
